Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-15 examples.tex
+

the appropriate L1 page table entry (Line \ref{line:l1entry_store} in Figure \ref{fig:mapping_code});

@@

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{walkpgdir} (discussed again in Section \ref{sec:traversing}).
\textsf{walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$)
both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
\looseness=-1
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\ghostmaptoken{\delta{}s}{\mathit{cr3val}}{\delta}) \ast 
  \fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac} \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

The crucial step is actually updating the L1 entry (Line \ref{line:l1entry_store} in Figure \ref{fig:mapping_code}),
via the virtual address
(\textsf{pt\_entry+KERNBASE}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$).
% , to hold the freshly
% allocated physical page address (\textsf{fpaddr}) in Line X.
After this write, we have
% \[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by the PTE variant of Rule}~\TirNameStyle{WriteToVirtMemFromReg})\]
\[\texttt{pt\_entry+KERNBASE} \mapsto_{\texttt{vpte}}  \; \paddr\; (\textsf{fpaddr \& 0x1})  \quad (\textsf{by PTE variant of}~\TirNameStyle{WriteToVirtMemFromReg})\]

At this point in the proof, % $\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; (\textsf{fpaddr})$ 
the virtual PTE points-to
contains  full ownership of the L1 table entry's backing physical memory.
Then, we can split the full-ownership of that
physical pointsto
% ($\textsf{fpaddr}\mapsto_{\textsf{p}} \;\textsf{wzero 64}$) 
into 512 pieces to obtain
fractional ownership of the physical memory appropriate for a single virtual address's share of an L1 entry.
% \todo{ismail comment out the rest of this subsection and fix according to the new map-a-page}
\begin{comment}
Since we know $\texttt{addr\_L1 }(\vaddr,
\entryo) \mapsto_{\mathsf{p}} \{q4\} \;\textsf{fpaddr}$) from
$\mapsto_{\textsf{vpte}}$, after the split of the PTE points-to, and, thanks to the post-condition of the page-table walk (Specification of \textsf{pte\_walkpgdir} in Figure \ref{}), we know the existence of L4-L2 mappings reaching to the L1 exist, so that we can add the last mapping $\texttt{fpaddr} \mapsto_{p} (\texttt{wzero 64})$ to obtain the complete table-walk.
\todo{with the virtual-pte-pointsto assertions (L4\_L1\_PointsTo($\vaddr$ l4e l3e l2e
fpaddr)) in Line 25}

Finally, we can insert $\vaddr$ into the ghost page-table-walk summarization
map ($\theta$) to change its state from unmapped ($\ulcorner \theta
\;!!\;\vaddr = \texttt{None}\urcorner$ (Line 28) and the
precondition) to mapped (Line 29) using Iris' ghost-map update
($\sqsubseteq$), and construct a PTE -points-to $\vaddr
\mapsto_{\textsf{vpte}}\; \qfrac \;\fpaddr\;(\textsf{wzero 64})$ (Line 31) using $\sumwalkabs\vaddr\qfrac\fpaddr$ obtained from ghost
page-table-walk insertions, $\ghostmaptoken{\delta{}s}{\rtv}{\delta}$ from unfolding the
definition of $\mapsto_{\textsf{vpte}}$, and $\fpaddr
\mapsto_{\textsf{p}} \textsf{wzero 64}$.
\end{comment}
Finally, by existentially quantifying away \textsf{fpaddr},
we can shift the PTE points-to into a normal virtual points-to.

For brevity this proof only maps the first word of the allocated page; the generalization to the full page
is straightforward use of iterated separating conjunction to manage collections of clusters of 512 things
(slices of the L1 entry physical ownership, virtual addresses at 8 byte increments, etc.).
\looseness=-1

>>>>

the appropriate L1 page table entry (Line 11 in Figure \ref{fig:mapping_code});

@@

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{walkpgdir} (discussed again in Section \ref{sec:traversing}).
\textsf{walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$)
both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
\looseness=-1
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\ghostmaptoken{\delta{}s}{\mathit{cr3val}}{\delta}) \ast 
  \fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac} \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

@@

The crucial and the only step in addition to traversing the page table in Figure \ref{walkpgdir} is actually updating the L1 entry (Line 11 in Figure \ref{fig:mapping_code}),
via the virtual address
(\textsf{pt\_entry+KERNBASE}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_to\_pfn(pa)}$).
% , to hold the freshly
% allocated physical page address (\textsf{fpaddr}) in Line X.

At this point in the proof, i.e. after the traversal to obtain the sound physical path resources ($\textsf{R}_{\textsf{walk\_acc}}$ Line 27 in Figure \ref{walkpgdir}), which coontains
virtual PTE points-to relations along the path to the full ownership of the L1 table entry's backing physical memory, we update the L1 entry to point the fresh page, as we do for the non-existent entries in the page table traversal (Line 6 in Figure \ref{pteinitializespec}.   
\looseness=-1

### 2024-11-15 intro.tex
+

\paragraph{Contributions}
We develop these ideas in the form of a logic for working with virtual-address-space-relative assertions,
implemented as an embedded separation logic within Iris~\cite{jung2018iris}.
The result is a separation logic that lifts a number of major semantic restrictions present in the few
prior logics tackling virtual address translation.
The logic we develop covers core reasoning principles for reasoning about memory configurations and code
reliant upon or manipulating those memory configurations in the presence of in-memory page tables, the primary
memory protection mechanism across Intel/AMD's x86-64 processors, ARM's application class processors including
AArch64 CPUs, POWER, RISC-V, and other architectures.
We prove the soundness of our \textsf{vProp} logic with respect to a RISC-like fragment of \textsf{AMD64} instructions.
We verify simplified versions of several critical virtual-memory-related pieces of OS functionality, 
including mapping  pages, switching address spaces, and converting physical addresses to virtual addresses for
software page table walks (as when mapping pages). These examples demonstrate the suitability and flexibility of
a modal treatment of address spaces: each has a concise specification, each example either goes beyond the technical
capabilities of prior logics, or revisits an example from prior work with more details (treating parts of the kernel
invariant never addressed by prior work) and fewer assumptions (e.g., verifying virtual-to-physical conversion rather
than assuming it has already occurred).

>>>>

\paragraph{Contributions}
We develop these ideas in the form of a logic for working with virtual-address-space-relative assertions,
implemented as an embedded separation logic within Iris~\cite{jung2018iris}.
The result is a separation logic that lifts a number of major semantic restrictions present in the few
prior logics tackling virtual address translation.
The logic we develop covers core reasoning principles for reasoning about memory configurations and code
reliant upon or manipulating those memory configurations in the presence of in-memory page tables, the primary
memory protection mechanism across Intel/AMD's x86-64 processors, ARM's application class processors including
AArch64 CPUs, POWER, RISC-V, and other architectures.
We prove the soundness of our \textsf{vProp} logic with respect to a RISC-like fragment of \textsf{AMD64} instructions.
We verify simplified versions of several critical virtual-memory-related pieces of OS functionality, most importantly:
switching address spaces, converting physical addresses to virtual addresses for
software page table walks, and complete page table walk which backbones other essential functionalities of
virtual-memory which we discuss for mapping and unmapping pages. These examples demonstrate the suitability and flexibility of
a modal treatment of address spaces: each has a concise specification, each example either goes beyond the technical
capabilities of prior logics, or revisits an example from prior work with more details (treating parts of the kernel
invariant never addressed by prior work) and fewer assumptions (e.g., verifying virtual-to-physical conversion rather
than assuming it has already occurred).
